---
title: "Multi-table transforms"
format:
  html:
    css: styles.css
---

![Image: *Merge* by Steve Walser is licensed under [CC BY-NC-ND 2.0](https://creativecommons.org/licenses/by-nc-nd/2.0/)](figures/data-shaping-4.png)

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

Using **base R**, **data.table**, and **dplyr** to transform MIDFIELD data as needed for your analysis.

- [Introduction](data-shaping-00-introduction.html)
- [Data selection](data-shaping-01-data-selection.html)
- [Basic transforms](data-shaping-02-basic-transforms.html)
- [Aggregating transforms](data-shaping-03-aggregating-transforms.html)
- [Multi-table transforms](data-shaping-04-multi-table-transforms.html)  &#9665; *You are here.*
- [Reshaping transforms](data-shaping-05-reshaping-transforms.html)

In this section, ...



## Getting started

Packages and datasets, as shown in the  [Introduction](data-shaping-00-introduction.html). 

```{r}
library(midfielddata) # datasets
library(data.table) # data manipulation system
library(dplyr) # data manipulation system
library(tidyr) # drop_na()

data(student, term)

student_base  <- data.frame(student)
student_DT    <- copy(student)
student_dplyr <- tibble(student)

term_base  <- data.frame(term)
term_DT    <- copy(term)
term_dplyr <- tibble(term)
```


## Appending rows

If two data frames have the same number of columns, with the same names, in the same order, one set of rows is easily appended to the other. 

Two data frames have been prepared for this task, saved in the `data` directory of the Institute website. Read them in as follows:

```{r}
# data URL
data_url <- "https://raw.githubusercontent.com/MIDFIELDR/2024-midfield-institute/main/data/ever_enrolled.csv"

# download the data
ever_enrolled <- fread(data_url)

# write the data to your data directory
fwrite(ever_enrolled, "data/ever_enrolled.csv")git add .
git commit -m "data-shaping-05"
```




## Splitting tables





## Appending columns 






## Left join





## Inner join 





## References 

<div id="refs"></div>

